---
title: "World Population in 1800"
format: dashboard
theme: lux
editor_options: 
  chunk_output_type: console
---


```{r}
# Load packages 
#if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse,
               plotly,
               here, 
               bslib, 
               bsicons, 
               shiny, 
               rgeoboundaries, 
               rnaturalearth,
               ggplot2,
               countrycode, 
               htmltools, 
               reactable)

```

```{r}
# Read in data
pop_raw <- 
  read_csv(here("data/pop.csv"))

# glimpse(pop_raw)
# 
# head(pop_raw)

pop_long <- 
  pop_raw %>% 
  pivot_longer(cols = `1800`:`2100`,
               names_to = "year",
               values_to = "values")

```

```{r}
# Convert values column to numeric
pop_long <- 
  pop_long %>%
  mutate(values = case_when(
    grepl("M$", values) ~ as.numeric(sub("M$", "", values)) * 1e6,
    grepl("k$", values) ~ as.numeric(sub("k$", "", values)) * 1e3,
    TRUE ~ as.numeric(values)
  ))


# Remove rows with NA values
 pop_long <- 
   pop_long %>% drop_na()
```


```{r}
# Custom function to format numbers
format_number <- function(x) {
  if (x >= 1e3 & x < 1e6) {
    return(paste0(round(x / 1e3, 1), "k"))
  } else if (x >= 1e6) {
    return(paste0(round(x / 1e6, 1), "M"))
  } else {
    return(as.character(x))
  }
}

# # Example usage
# numbers <- c(999, 1000, 2500, 500000, 6000000)
# formatted_numbers <- sapply(numbers, format_number)
# print(formatted_numbers)
```


```{r}
# Extract countries and values
lowest_population_1800_country <- 
  pop_long %>% 
  filter(year == 1800) %>% 
  filter(values == min(values)) %>% 
  pull(country)

lowest_population_1800_population <- 
  pop_long %>% 
  filter(year == 1800) %>% 
  filter(values == min(values)) %>% 
  pull(values)


# Filter the data for the year 2010
pop_1800 <- 
  pop_long %>% filter(year == 1800)

# Calculate the median value
median_value_1800 <- 
  median(pop_1800$values)

# Filter the data to get the row(s) with the median value
median_population_1800_country <- 
  pop_1800 %>% 
  filter(values == median_value_1800) %>% 
  pull(country)

median_population_1800_population <- 
  pop_1800 %>% 
  filter(values == median_value_1800) %>% 
  pull(values)
  
median_population_1800_population <-
  format_number(median_population_1800_population)


highest_population_1800_country <- 
  pop_long %>% 
  filter(year == 1800) %>% 
  filter(values == max(values)) %>% 
  pull(country)

highest_population_1800_population <- 
  pop_long %>% 
  filter(year == 1800) %>% 
  filter(values == max(values)) %>% 
  pull(values)

highest_population_1800_population <-
  format_number(highest_population_1800_population)

```

```{r}
country_shapes <- 
  rnaturalearth::ne_countries()

# ggplot(country_shapes) +
#   geom_sf()

pop_1800 <- 
  pop_long %>% 
  filter(year == 1800)

pop_1800 <- 
pop_1800 %>% 
  mutate(
    country_code = countrycode(country, 
                    "country.name", 
                    "iso3c"))
# pop_long
# 
# setdiff(pop_long$country_code, country_shapes$adm0_a3)

gap_map <- 
  left_join(country_shapes,
          pop_1800,
          by = c("adm0_a3" = "country_code")) %>% 
  mutate(tooltip_label = paste(country,
                               round(values,1),
                               sep = ": "
                               )) %>% 
  ggplot() +
  geom_sf(aes(fill = values,
              text = tooltip_label)) +
  theme_void() +
  scale_fill_continuous(trans = "reverse") +
  theme(legend.position = "none")

gap_map_ggplotly <- 
  ggplotly(gap_map,
         tooltip = "text")
```


# HOME

## Row 1 {height=25%}

```{r}
value_box(
  title = "Lowest Population in 1800",
  value = lowest_population_1800_population,
  showcase = bsicons::bs_icon("heart"),
  theme = value_box_theme(bg = "#008080"),
  p(paste0("(", lowest_population_1800_country, ")"))
)
```


```{r}
value_box(
  title = "Median Population in 1800",
  value = median_population_1800_population,
  showcase = bsicons::bs_icon("thermometer"),
  theme = value_box_theme(bg = "#008080"),
  p(paste0("(",median_population_1800_country, ")"))
)
```


```{r}
value_box(
  title = "Highest Population in 1800",
  value = highest_population_1800_population,
  showcase = bsicons::bs_icon("thermometer"),
  theme = value_box_theme(bg = "#008080"),
  p(paste0("(",highest_population_1800_country, ")"))
)
```


## Row 2 {height=75%}


### {width=60%}


```{r title = "Map of Countries Population in 1800"}
gap_map_ggplotly
```

```{r}
top_15_pop_1800 <- 
  pop_1800 %>% 
  arrange(desc(values)) %>% 
  head(15) %>%
  mutate(tooltip_label = paste(country,
                               round(values,1),
                               sep = ": "
                               )) %>% 
  ggplot(aes(y = reorder(country, values),
             x = values,
             fill = values,
             text = tooltip_label)) +
  geom_col()+
  geom_text(aes(label = round(values, 1)),
            nudge_x = -10,
            color = "black") +
  labs(y = "Country",
       x = "Population") +
  scale_fill_continuous(trans = "reverse") +
  theme(legend.position = "none")


top_15_pop_1800_ggplotly <- 
  ggplotly(top_15_pop_1800,
           tooltip = "text")


```


### {width=40%}


```{r title = "Top 15 Countries by Population in 1800"}
top_15_pop_1800_ggplotly
```




# DOWNLOAD DATA

The data used in this dashboard is shown below and can be downloaded as a CSV.

```{r}

htmltools::browsable(
  tagList(
    reactable(pop_raw,
              elementId = "population-table",
              searchable = T,
              filterable = T),
  
    tags$button("Download as CSV", 
                onclick = 
                  "Reactable.downloadDataCSV('population-table')")
  )
)


```


# ABOUT

This data comes from the r `gapminder` package and is originally sourced from the Gapminder Foundation.

Gapminder Foundation is a non-profit venture registered in Stockholm, Sweden, that promotes sustainable global development and achievement of the United Nations Millennium Development Goals by increased use and understanding of statistics and other information about social, economic, and environmental development at local, national, and global levels.

Gapminder was founded in 2005 by Ola Rosling, Anna Rosling Rönnlund, and Hans Rosling. The name Gapminder was derived from the "Mind the Gap" warning messages on the London Underground.

An example of Hans Rosling's videos is shown below:

<iframe width="560" height="315" src="https://www.youtube.com/embed/hVimVzgtD6w?si=j0LDxdS-EUQHAnRh" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>